In this proposal, we aim to formalize a subset of \llvm \ir
and a selection of peephole optimizations in \lean~\cite{moura2021lean},
a proof assistant.
\cref{tab:formalization-scope} lists the scope of the proposed formalization.

Compared with existing projects like \vellvm~\cite{zhao2012formalizing} 
and \leanmlir~\cite{bhat2024verifying}, we focus on 
a small subset of \llvm \ir and a small selection of peephole optimizations,
to facilitate my learning process.

\begin{table}[ht]
\centering
\footnotesize
\setlength{\tabcolsep}{4pt}\renewcommand{\arraystretch}{1}
\caption{Scope of the Proposed Formalization in \lean}
\label{tab:formalization-scope}
\begin{tabular}{@{}l p{0.8\linewidth}@{}}
\toprule
\textbf{Category} & \textbf{Target functionalities and concepts} \\
\midrule
Instructions & \texttt{add}, \texttt{sub}, \texttt{mul}, \texttt{div},
\texttt{select}, \texttt{load}, \texttt{store}; flags: \texttt{nsw},
\texttt{nuw} \\
Data types & Arbitrary-width integers (\eg \texttt{i32}, \texttt{i64}),
pointer types; special values: \poison, \vundef \\
Memory model & Basic locations/addresses; operational semantics of
\texttt{load}/\texttt{store}; no aliasing or complex hierarchies \\
Optimizations & Arithmetic identities (e.g. \texttt{add x,0} $\to$
\texttt{x}); constant folding (e.g. \texttt{select true,x,y} $\to$
\texttt{x}) \\
Properties & Refinement: $P_{opt}\sqsubseteq P_{orig}$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Instructions}

We focus on a subset of \llvm \ir with integer arithmetic and memory
operations that are typical targets for peephole transforms. This subset
includes \texttt{add}, \texttt{sub}, \texttt{mul}, \texttt{div},
\texttt{select}, \texttt{load}, and \texttt{store}, together with
relevant flags (\eg \texttt{nuw}, \texttt{nsw}) that affect semantics.

\subsection{Data Types}

We formalize integer types of varying widths (\eg, \texttt{i32},
\texttt{i64}) and pointer types, and model \poison and \vundef to capture
\llvm's semantics relevant to optimization correctness.

\subsection{Memory Model}

The memory model captures basic locations, addresses, and the operational
effects of \texttt{load} and \texttt{store}. We omit aliasing and complex
hierarchies in this initial formalization.


\subsection{Peephole Optimizations}

We formalize common peephole optimizations that target arithmetic and
memory patterns, e.g. \texttt{add x,0} $\rightarrow$ \texttt{x} and
constant folding of \texttt{select} when the condition is constant.

\subsection{Properties to Verify}

For each optimization we prove a refinement property: the optimized
program refines the original ($P_{opt}\sqsubseteq P_{orig}$), i.e. it is
semantically equivalent or more defined where the original may yield
\poison{} or \vundef.

