\section{Lessons Learned}
\label{sec:lessons}

This project clarified a few points that were not obvious at the proposal
stage. The main lessons concern where the real complexity lives, which design
choices most improved proof development, and how scope control affected the
quality of the final artifact.

\myparagraph{Semantic Modeling}
The first lesson is that semantic modeling dominates everything else. Before
this project, we expected the interesting work to be mostly in proving algebraic
rewrites. In practice, the deeper work was deciding what counts as a program,
what counts as a well-formed input, what exactly execution returns, and how
refinement should quantify over semantic effects such as \vundef. Once these
choices were made well, many individual proofs became much easier.

\myparagraph{Executable Semantics}
The second lesson is that executable semantics is a powerful proving aid.
Because the semantics reduces by computation, small examples can often be
proved by simplification after the surrounding proof obligations are arranged
correctly. This reinforced an appealing idea about proof assistants: one does
not always have to reason \emph{about} an interpreter abstractly if one can
also reason \emph{with} the interpreter computationally.

\myparagraph{Surface Syntax}
The third lesson is that surface syntax matters even in a proof project.
Readable theorem statements changed our willingness to experiment. It was much
easier to explore optimization ideas when we could write something that looked
like \llvm rather than deeply nested constructors.

\myparagraph{Scope Control}
The fourth lesson is more personal. We learned that there is value in
narrowing scope early and doing so explicitly. The proposal was useful because
it forced us to think ambitiously, but the implemented project improved once we
accepted that finishing memory, optimizer infrastructure, and proofs all at
once would likely produce a weaker result. Focusing on a smaller but tighter
artifact was not merely damage control; it was a real design improvement.
